---
title: "STA2201 Lab 1"
author: "Timothy Jordan Regis: 1005296323"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
```

# Data

```{r, include=FALSE}
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
head(dm)
dm <- data.frame(dm)

pop <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", skip = 2, col_types = "dcddd")
head(pop)
pop <- data.frame(pop)
```
Mortality Rate Data:

```{r}
head(dm)
```

Population Data:

```{r}
head(pop)
```


# Q1
```{r}
mort_rates = dm %>% 
  filter(
  Age == 10 | Age == 20 | Age == 30 | Age == 40 
) %>% 
  mutate(
  ratio = Male/Female
)
mort_rates %>% 
  ggplot(aes(y = ratio, x = Year, color = Age)) +
  geom_line(size = 1) + 
  labs(title = "Male/Female Mortality Rates") +
  ylab("Ratio") + 
  theme_minimal()
#plot(mort_rates$ratio, mort_rates$Year, type = "line", color = Age)
```

# Q2

```{r}
max_females = dm %>% group_by(Year) %>% filter(Female == max(Female, na.rm = TRUE)) %>% select(Year, Age, Female) %>% rename("Rate" = Female)
head(max_females)
```

```{r}
# length(unique(dm$Year))
# length(unique(max_females$Year))
```

```{r}
kable(max_females, digits = 3, booktabs = FALSE, longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "repeat_header"))
```



# Q3



```{r, message=FALSE, warning=FALSE}
stds = dm %>% 
  group_by(Age) %>% 
  summarise(std_Male = sd(Male, na.rm = TRUE), 
            std_Female = sd(Female, na.rm = TRUE), 
            std_Total = sd(Total, na.rm = TRUE))
stds[order(as.numeric(stds$Age)),]
```


```{r}
#length(unique(dm$Age))
```


```{r, message=FALSE, warning=FALSE}
kable(stds[order(as.numeric(stds$Age)),], digits = 5, booktabs = TRUE, longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "repeat_header"))
```



# Q4

```{r}
m_vs_f = dm %>% left_join(y = pop, by = c("Year", "Age")) %>% mutate(
  pw_F = Female.x/Female.y,
  pw_M = Male.x/Male.y
) %>% filter(!is.nan(pw_M) & !is.nan(pw_F) & 
             !is.infinite(pw_M) & !is.infinite(pw_F)) %>% 
  group_by(Year) %>% 
  summarise(avg_F = mean(pw_F, na.rm = TRUE), avg_M = mean(pw_M, na.rm = TRUE)) %>% 
  pivot_longer(cols = c(2:3), names_to = "Sex", values_to = "Average_Rate") 

m_vs_f %>% 
  ggplot(aes(x = Year, y = Average_Rate, color = Sex)) + 
    geom_line(size = 1) + 
    labs(title = "Male vs. Female Death Rates") +
    ylab("Average Population Weighted Death Rate") +
    scale_color_manual(labels = c("Female", "Male"), values = c("darkorchid", "forestgreen")) + 
  theme_minimal()
  # + 
  #   scale_fill_discrete(labels=c('Female', 'Male'))
```

From the looks of this graph, men tend to see a much higher variability in their death rates, often attaining greater average levels than Women.
This is especially notable in the years after the 1960s, where men's rates fluctuate over a very short period while women's remain fairly constant. 
between the women's rates and rates much greater.






