---
title: "STA2201 Lab #6"
author: "Timothy Jordan Regis"
date: "23/02/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Introduction

This lab will be looking at trying to replicate some of the visualizations in the lecture notes, involving prior and posterior predictive checks, and LOO model comparisons. 

The dataset is a 0.1% of all births in the US in 2017. I've pulled out a few different variables, but as in the lecture, we'll just focus on birth weight and gestational age. 

# The data

Read it in, along with all our packages. 

```{r, include = FALSE}
library(tidyverse)
library(here)
library(ggpubr)
library(ggcorrplot)
# for bayes stuff
library(rstan)
library(bayesplot) 
library(loo) 
library(tidybayes) 
library(readr)

ds <- read_rds(here("~/Grad/STA2201/lab6mak/","births_2017_sample.RDS"))
```


```{r}
head(ds)
```

Brief overview of variables:

- `mager` mum's age
- `mracehisp` mum's race/ethnicity see here for codes: https://data.nber.org/natality/2017/natl2017.pdf page 15
- `meduc` mum's education see here for codes: https://data.nber.org/natality/2017/natl2017.pdf page 16
- `bmi` mum's bmi 
- `sex` baby's sex
- `combgest` gestational age in weeks
- `dbwt` birth weight in kg
- `ilive` alive at time of report y/n/ unsure

I'm going to rename some variables, remove any observations with missing gestational age or birth weight, restrict just to babies that were alive, and make a preterm variable. 

```{r}
ds <- ds %>% 
  rename(birthweight = dbwt, gest = combgest) %>% 
  mutate(preterm = ifelse(gest<32, "Y", "N")) %>% 
  filter(ilive=="Y",gest< 99, birthweight<9.999)
```


## Question 1

Use plots or tables to show three interesting observations about the data. Remember:

- Explain what your graph/ tables show
- Choose a graph type that's appropriate to the data type
- If you use `geom_smooth`, please also plot the underlying data

Feel free to replicate one of the scatter plots in the lectures as one of the interesting observations, as those form the basis of our models. 

```{r}
ds %>%
  ggplot(aes(log(gest), log(birthweight), color = preterm)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = TRUE) + 
  scale_color_brewer(palette = "Set1") + 
  theme_bw(base_size = 14) + 
  ggtitle("Birthweight vs. Gestational Age by Preterm Status") +
  xlab("Gestational Age (Logs)") + 
  ylab("Birthweight (Logs)") +
  theme(plot.title = element_text(hjust = 0.5))
```

In our first plot, we see the distribution of birthweights by gestational age, both on the log scale, with a colouring by the preterm status. We have also added linear smoothings to each of these groups to better identify the existing trends.
As we can see, there is a fairly clear positive relationship between gestational age and birth weight, suggesting longer gestation peiods leading to heavier babies, possibly due to increased nutrition in the womb. Furthermore, there is a difference of effects based on the preterm status, where we see preterm babies having a much stronger positive relationship than babies born after the preterm period.


```{r}
ds %>%
  ggplot(aes(birthweight, color = preterm)) + 
  geom_density() + 
  #geom_smooth(method = "lm") + 
  scale_color_brewer(palette = "Set1") + 
  theme_bw(base_size = 14) + 
  ggtitle("Birthweight Distribution by Preterm Status") +
  xlab("Birthweight") + 
  ylab("Density") +
  theme(plot.title = element_text(hjust = 0.5))
```

Our second plot displays the distribution of birthweights in the data, grouped again by the preterm status. Here, we can observe that there is a significant difference between the birthweights of preterm and non-preterm babies, where preterm births have a high density at lower weights, peaking at 1, whereras non-preterm babies have a majority of their density at weights above 2, peaking around 3.5. Thus, we can see that not only do other factors have different effects based on preterm status, but that the birthweights themselves are also potentially influenced by the preterm status.
This result is faairly sensible, however, as we acknowledge the multitude of difficulties presented in preterm births, and the much greater frequency of complications that arise which can impact the weight of the child. Moreover, preterm babies spend less time in the womb and thus have less time to grow in the final weeks of pregnancy which are often considered to be the most valuable to a baby's development.


```{r}
ggcorrplot(cor(ds %>% mutate(sex = as.numeric(as.factor(sex)), preterm = as.numeric(as.factor(preterm)), ilive = as.numeric(as.factor(ilive))) %>% select(c(1:7, 9, 8))), lab = TRUE) + ggtitle("Correlation Matrix")
```

Our final plot displays the correlation matrix of our data, which can aid in identifying any interesting links between the raw data presented.
As we can see firstly, the column tracking the survival of a baby has no correlation due to our filtering out all deaths. We can also notice strong correlations between a few variables. Firstly, preterm status has a strong negative correlation with both birthweight and gestational period, which agree with the observations we saw in the previous two plots. Furthermore, birthweight has a high positive correlation with gestational age, again agreeing with our previous plots. The education of a mother also has strong correlations with their age and race. With age, we see a positive relationship, which suggests that more educated mothers are often correlated with older mothers. Lastly, with race, this becomes somewhat uninterpretable, as unlike education, which has ordinal characteristics with higher numbers equalling higher education levels, race is purely categorical, so we cannot directly interpret how these variables are related exactly.
The remaining variables showed no significant correlations above a magnitude of 0.2.


# The model

As in lecture, we will look at two candidate models 

Model 1 has log birth weight as a function of log gestational age

$$
\log(y_i) \sim N(\beta_1 + \beta_2\log(x_i), \sigma^2)
$$

Model 2 has an interaction term between gestation and prematurity

$$
\log(y_i) \sim N(\beta_1 + \beta_2\log(x_i) + \beta_3 z_i + \beta_4\log(x_i) z_i, \sigma^2)
$$

- $y_i$ is weight in kg
- $x_i$ is gestational age in weeks, CENTERED AND STANDARDIZED
- $z_i$ is preterm (0 or 1, if gestational age is less than 32 weeks)


# Prior predictive checks

Let's put some weakly informative priors on all parameters i.e. for the $\beta$s

$$
\beta \sim N(0, 1)
$$

and for $\sigma$

$$
\sigma \sim N^+(0,1)
$$
where the plus means positive values only i.e. Half Normal. 

Let's check to see what the resulting distribution of birth weights look like given Model 1 and the priors specified above, assuming we had no data on birth weight (but observations of gestational age).

## Question 2

For Model 1, simulate values of $\beta$s and $\sigma$ based on the priors above. Do 1000 simulations. Use these values to simulate (log) birth weights from the likelihood specified in Model 1, based on the set of observed gestational weights. **Remember the gestational weights should be centered and standardized**. 

- Plot the resulting distribution of simulated (log) birth weights. 
- Plot ten simulations of (log) birthweights against gestational age. 


# Run the model

Now we're going to run Model 1 in Stan. The stan code is in the `code/models` folder. 

First, get our data into right form for input into stan. 

```{r}
ds$log_weight <- log(ds$birthweight)
ds$log_gest_std <- (log(ds$gest) - mean(log(ds$gest)))/sd(log(ds$gest))

# put into a list
stan_data <- list(N = nrow(ds),
                  log_weight = ds$log_weight,
                  log_gest = ds$log_gest_std)
```

Now fit the model

```{r, include = FALSE}
mod1 <- stan(data = stan_data, 
             file = here("~/Grad/STA2201/lab6mak/simple_weight.stan"),
             iter = 500,
             seed = 243)
```

```{r}
summary(mod1)$summary[c("beta[1]", "beta[2]", "sigma"),]
```

```{r, message = FALSE, warning=FALSE}
set.seed(182)
nsims <- 1000
sigma <- abs(rnorm(nsims, 0, 1)) #1 / sqrt(rgamma(nsims, 1, rate = 100))
beta0 <- rnorm(nsims, 0, 1)
beta1 <- rnorm(nsims, 0, 1)

dsims <- tibble(log_gest_std = (log(ds$gest)-mean(log(ds$gest)))/sd(log(ds$gest)))

for(i in 1:nsims){
  this_mu <- beta0[i] + beta1[i]*dsims$log_gest_std 
  dsims[paste0(i)] <- this_mu + rnorm(nrow(dsims), 0, sigma[i])
}

dsl <- dsims %>% 
  pivot_longer(`1`:`10`, names_to = "sim", values_to = "sim_weight")

dsl %>% 
  ggplot(aes(sim_weight)) + 
  geom_histogram(aes(y = ..density..), bins = 50, fill = "turquoise", color = "black") + 
  ggtitle("Simulated Log Birthweights") + 
  theme_bw(base_size = 16)
  #xlim(c(-1000, 1000)) + 
  #geom_vline(xintercept = log(60), color = "purple", lwd = 1.2, lty = 2) + 
  


dsims_samps <- dsims[, sample(ncol(dsl), 10, replace = FALSE)]
colnames(dsims_samps) <- paste("Sim", 1:10, sep = "_")
dsims_samps <- dsims_samps %>%
  mutate(log_gest_std = dsims$log_gest_std)
dsims_samps_l <- dsims_samps %>%
  pivot_longer(cols = c(1:10), names_to = "Simulation", values_to = "Sim_weight")

dsims_samps_l %>%
  ggplot(aes(x = log_gest_std, y = Sim_weight, color = Simulation)) +
  geom_point() +
  #geom_smooth(method = "lm") +
  ggtitle("Log Birthweights vs. Log Gest. Age") + 
  xlab("Log Gestational Age (Centered)") +
  theme_bw(base_size = 16) +
  theme(plot.title = element_text(hjust = 0.5))
```

Both plots display a centering of data and observations around 0 for weight which was our initial mean value. No significant issues observed.

## Question 3

Based on model 3, give an estimate of the expected birthweight of a baby who was born at a gestational age of 37 weeks. 

```{r}
summary(mod1)$summary[c("beta[1]", "beta[2]", "sigma"),]
```
```{r, echo = TRUE}
exp(1.1626250 + ((log(37)-mean(log(ds$gest)))/sd(log(ds$gest)))*0.1436183)
```

Thus, we see that the expected birthweight of a baby born at 37 weeks is approximately 2.94


## Question 4

Write a stan model to run Model 2, and run it. 


```{r}
ds2 <- ds %>% 
  mutate(z = case_when(preterm == "Y" ~ 1, 
                              TRUE ~ 0))
ds2$log_weight <- log(ds2$birthweight)
ds2$log_gest_std <- (log(ds2$gest) - mean(log(ds2$gest)))/sd(log(ds2$gest))

# put into a list
stan_data2 <- list(N = nrow(ds2),
                  log_weight = ds2$log_weight,
                  log_gest = ds2$log_gest_std,
                  z = ds2$z)
```

Now fit the model

```{r, include = FALSE}
mod_2 <- stan(data = stan_data2, 
             file = here("~/Grad/STA2201/lab6mak/simple_weight_model2.stan"),
             iter = 500,
             seed = 243)
```


```{r}
summary(mod_2)$summary[c(paste0("beta[", 1:4, "]"), "sigma"),]
```


## Question 5

For reference I have uploaded some model 2 results. Check your results are similar. 

```{r}
load(here("~/Grad/STA2201/lab6mak/", "mod2out.Rda"))
summary(mod2)$summary[c(paste0("beta[", 1:4, "]"), "sigma"),]
```

Coefficients are quite similar with minor differences in sd and se.



# PPCs

Now we've run two candidate models let's do some posterior predictive checks. The `bayesplot` package has a lot of inbuilt graphing functions to do this. For example, let's plot the distribution of our data (y) against 100 different datasets drawn from the posterior predictive distribution:

```{r}
set.seed(1856)
y <- ds$log_weight
yrep1 <- extract(mod1)[["log_weight_rep"]]
yrep2 <- extract(mod_2)[["log_weight_rep"]] 
dim(yrep1)
samp100 <- sample(nrow(yrep1), 100)
ppc_dens_overlay(y, yrep1[samp100, ])  + ggtitle("distribution of observed versus predicted birthweights")
```

## Question 6

Make a similar plot to the one above but for model 2, and **not** using the bayes plot in built function (i.e. do it yourself just with `geom_density`)

```{r}
samp100 <- sample(nrow(yrep2), 100)

# first, get into a tibble
rownames(yrep2) <- 1:nrow(yrep2)
dr <- as_tibble(t(yrep2))
dr <- dr %>% bind_cols(i = 1:nrow(ds), log_weight_obs = log(ds$birthweight))

# turn into long format; easier to plot
dr <- dr %>% 
  pivot_longer(-(i:log_weight_obs), names_to = "sim", values_to ="y_rep")

# filter to just include 100 draws and plot!
dr %>% 
  filter(sim %in% samp100) %>% 
  ggplot(aes(y_rep, group = sim)) + 
  geom_density(alpha = 0.2, aes(color = "y_rep")) + 
  geom_density(data = ds %>% mutate(sim = 1), 
               aes(x = log(birthweight), col = "y")) + 
  scale_color_manual(name = "", 
                     values = c("y" = "darkblue", 
                                "y_rep" = "lightblue")) + 
  ggtitle("Dist. of Observed and Replicated Birthweights") + 
  theme_bw(base_size = 16)
```

## Test statistics

We can also look at some summary statistics in the PPD versus the data, again either using `bayesplot` -- the function of interest is `ppc_stat` or `ppc_stat_grouped` -- or just doing it ourselves using ggplot. 

E.g. medians by prematurity for Model 1

```{r}
ppc_stat_grouped(ds$log_weight, yrep1, group = ds$preterm, stat = 'median')
```

## Question 7

Use a test statistic of the proportion of births under 2.5kg. Calculate the test statistic for the data, and the posterior predictive samples for both models, and plot the comparison (one plot per model).

```{r}
t_y <- mean(y<=log(2.5))
t_y_rep <- sapply(1:nrow(yrep1), function(i) mean(yrep1[i,]<=log(2.5)))
t_y_rep_2 <- sapply(1:nrow(yrep2), function(i) mean(yrep2[i,]<=log(2.5)))

ggplot(data = as_tibble(t_y_rep), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 1: Proportion of births less than 2.5kg") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue"))

ggplot(data = as_tibble(t_y_rep_2), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 2: Proportion of births less than 2.5kg") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue"))
```

As we can see, Model 2 has a higher proportion of its values around the observed value, suggesting an improvement




## Question 8

Based on the original dataset, choose one (or more) additional covariates to add to the linear regression model. Run the model in Stan, and compare with Model 2 above on at least 2 posterior predictive checks.

For our model we have chosen to add in the mother's age, on the log scale, and we first standardize this value

```{r}
ds3 <- ds %>%
  mutate(z = ifelse(ds$preterm=="Y",1,0),
         log_age_std = (log(mager) - mean(log(mager)))/sd(log(mager)))

# put into a list
stan_data3 <- list(N = nrow(ds3),
                  log_weight = ds3$log_weight,
                  log_gest = ds3$log_gest_std,
                  z = ds3$z,
                  log_age = ds3$log_age_std)
```

```{r, include = FALSE}
mod_3 <- stan(data = stan_data3, 
             file = here("~/Grad/STA2201/lab6mak/simple_weight_model3.stan"),
             iter = 500,
             seed = 243)
```


```{r, echo = TRUE}
# Given Model
summary(mod2)$summary[c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "sigma"),]
summary(mod_2)$summary[c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "sigma"),]
summary(mod_3)$summary[c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "beta[5]", "sigma"),]
```


```{r}
set.seed(1856)
#y <- log_weight

yrep2 <- extract(mod_2)[["log_weight_rep"]]
samp100_2 <- sample(nrow(yrep1), 100)
ppc_dens_overlay(y, yrep2[samp100_2, ]) + 
  ggtitle("Model 2: Dist. of Observed and Replicated Birthweights") + 
  ylab("Density") + 
  xlab("y_rep")

yrep3 <- extract(mod_3)[["log_weight_rep"]]
samp100_3 <- sample(nrow(yrep1), 100)
ppc_dens_overlay(y, yrep3[samp100_3, ]) + 
  ggtitle("Model 3: Dist. of Observed and Replicated Birthweights") + 
  ylab("Density") + 
  xlab("y_rep")

```

As we can see, there is very little difference between the two models, with both fitting quite close to the desired curve.

We can also observe the test statistic plots of each model.
```{r}
t_y <- mean(y<=log(2.5))
t_y_rep <- sapply(1:nrow(yrep2), function(i) mean(yrep2[i,]<=log(2.5)))
t_y_rep_2 <- sapply(1:nrow(yrep3), function(i) mean(yrep3[i,]<=log(2.5)))

ggplot(data = as_tibble(t_y_rep), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 2: proportion of births less than 2.5kg") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue"))

ggplot(data = as_tibble(t_y_rep_2), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 3: proportion of births less than 2.5kg") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue"))
```

Again, however, we still see quite similar results with no clear winner.


```{r}
loglik1 <- extract(mod_2)[["log_lik"]]
loglik2 <- extract(mod_3)[["log_lik"]]
loo1 <- loo(loglik1, save_psis = TRUE)
loo2 <- loo(loglik2, save_psis = TRUE)
```

```{r}
loo1
loo2
```

```{r}
#| echo: false
loo_compare(loo1, loo2)
```

Finally, to numerically assess the results, we can compare the two models via `loo_compare` which tells us that the new model, mod_3, is a little bit better than the earlier mod_2.


Note:

Helper code for this lab taken from: https://www.monicaalexander.com/posts/2020-28-02-bayes_viz/



